A Component-Based Sensor Network for \\ Environmental Monitoring

### 2009-03-13 docs/icccn2009/paper.tex
Edited NetBEAMS section
git-svn-id: svn://localhost/netbeams/trunk@375 559edde2-a4d4-44e0-96fc-4035e40d7396

The Gumstix runs Linux 2.6 with the BusyBox utilities, and uses the
OpenEmbedded cross-compile build environment to provide a complete
Linux environment and a large range of Linux applications for the ARM
architecture.  We use the JamVM \cite{jamvm01} implementation of Sun
Microsystem's virtual machine in order to run the DSP. The Gumstix
communicates with the YSI Sonde via an RS232 serial port using a
null-modem cable since both the Sonde and the Gumstix are DTE devices.
The serial port configuration is:

\begin{table}[!htb]
\caption{\label{tab:RS232_Config}RS232 Serial Port Configuration}
\centering
\begin{tabular}{l l}
\hline
Baud Rate&9600\\
Data Bit&8\\
Parity&None\\
Handshake&None\\
\hline
\end{tabular}
\end{table}

\begin{table}
\caption{\label{TAB_RS232_Pinout}RS232 Serial Pinout}
\centering
\begin{tabular} {|l|l|l|}
\hline
Wire Color	&Pin Description        &DB-9\\ 
\hline
Yellow 		&RS232 TX    &2\\
Orange		&RS232 RX    &3\\  
Green  		&Alarm       &----\\ 
Grey			&RTS         &----\\
Blue  		&CTS         &----\\
Red			&+ 12V DC    &9\\
Black			&GND         &5\\  
Purple		&SDI-12      &----\\ 
Bare			&Shield      &----\\
\hline
\end{tabular}
\end{table}

We make use of the cellular phone network to establish a communication
link with the backend server. The cellular phone network has the
advantages of reaching well beyond telemetry options such as wireless
networks and being a significantly lower cost solution than satellite
communications. We make use of the Huawei E220. The E220 is a USB
modem that supports HSDPA, UMTS, and EDGE packet data services at maximum
transmission rates of 3.6Mbps, 384kbps, and 236.8kbps respectively.
HSDPA and UMTS operate at 2100MHz while GSM, GPRS, and EDGE operate at
900, 1800, and 1900MHz. The modem connects to the Verdex Pro XM4
motherboard via a mini USB interface (supporting USB 2.0). Linux
kernels since 2.6.20 include support for E220 drivers.  This includes
the linux kernel on the Gumstix which runs 2.6.21.  The Gumstix
communicates with the modem using the PPP protocol and accesses the
cellular network using a data plan from AT\&T.

@@

Our implementation of the DSP is based on the Open Source
implementation of OSGi called Knopflerfish \cite{knopflerfish01}.
Since Sun Microsystems does not support its JDK for the Gumstix, we
use JamVM \cite{jamvm01} as the Java virtual machine implementation
and GNU Classpath \cite{classpath01} for the runtime libraries. We
have implemented various DSP Components that implement the
functionality needed for NetBEAMS. On the far left of Figure
\ref{FIG_NETBEAMS} is the actual sensor; in our case the
aforementioned YSI Sonde. A YSI Sonde data producer (1) is a DSP
component that uses the RS232 standard to read the physical
measurements from the device over a serial port. This component
converts the physical measurements into a DSP-specific message. The
following XML Schema on figure \ref{FIG_DSP_YSI_DATA} defines the YSI
Sonde's specific message format which includes each measurement data
point, described in Table \ref{TAB_SONDE_MEASUREMENTS}, and its type:

\begin{figure}[!htb]
\centering
\epsfig{file=dsp-message-body-ysi, width=8cm, height=7cm}
\caption{\label{FIG_DSP_YSI_DATA} Structure of the YSI Component Data.}
\end{figure}

The schema describes two Java classes, sondeDataType and
SondeDataMessage, which are used to package the Sonde's incoming data
into a DSP message for delivery to a remote consumer.  The
SondeDataMessage class extends the AbstractMessageContent class by
adding a new type to the message payload, SondeDataType.  Each time
the Sonde transmits its payload of measurement data during a sampling
interval, the data is converted into a SondeDataType which is then
encapsulated within SondeDataMessage and then sent to a remote
consumer.

The YSI Sonde component can also be configured via special DSP
messages, to direct the Sonde to change its sampling frequency.  These
property messages originate from the web frontend that we will
explained in a subsequent section.

@@

After producing the SondeDataContainer data, the YSI Sonde component
will wrap it up in a DSP Message Message and pass it to the DSP Broker
for its delivery.  However, our use case scenario is that the message
must go from the local sensor device to the server. For this reason,
we have developed a pair of client and server DSP components
specialized in transferring and receiving serialized objects throught
the communication channel using the HTTP protocol \cite{RFC2068}.

\begin{itemize}
\item DSP Wire Transport Client: takes a collection of DSP messages
  and transmit them using an HTTP client API written in JAVA
  \cite{http-client}. As a non-functional requirement, a DSP
  administrator can set the interval in which this component will
  transmit messages to other components;
\item DSP Wire Transport Server: is implemented using the Java
  Servlets technology \cite{java-servlets}, and is used to receive the
  collection of DSP messages transmitted, and potentially send DSP
  messages back to the sensor station.
\end{itemize}

One of the non-functional requirements for the DSP is to maximize the
utilization of the communication channel. For this specific reason, we
opted to add the type Messages Container, which contains a collection
of DSP messages, and specific identifiers of the hosting sensor.
Figure \ref{FIG_DSP_MSGS_CONTAINER} contains the structure of the DSP
Messages Container.

\begin{figure}[!htb]
\centering
\epsfig{file=dsp-messages-container, width=8cm}
\caption{\label{FIG_DSP_MSGS_CONTAINER} The structure of Messages Container.}
\end{figure}

Some remarks regarding the Messages Container and it's implementation are as
follows:

\begin{itemize}
  \item Container with or without Messages Container: an instance of
  MessagesContainer may or may not have DSP messages;
  \item Container may be used to acknowledge messages using a similar strategy
  as the Slide Windowing protocol \cite{slide-window}. 
\end{itemize}

\subsection{Sending Messages Container}

It's important to note that, at this point, the DSP Broker had already
checked with the Matcher which are the DSP components that will be
receive a copy of the message in question. The matcher had specific
instructions to use the DSP Wire Transport Client component as the
default ``Proxy'' or default gateway, if you prefer.

By building a MessagesContainer instance, the Wire Transport Client
adds all the messages addressed to the ``destinationHost'', with
specific UUID of the component (2).

\subsection{Receiving Messages Container}

After the DSP Wire Transport Server receives the Messages Container
(3), it unmarshalls the messages and forward them to the local DSP
Broker.

The reception of those messages will ALWAYS depend on the local DSP Broker and
its combination with the Matcher. The steps (4), (5) and (6) are the
actual reception of DSP messages. Container might also include an attribute
called acknowledgeUntil, which determines the next actions that will occur in the system.

>>>>

The Gumstix runs Linux 2.6 with the BusyBox utilities, and uses the
OpenEmbedded cross-compile build environment to provide a complete
Linux environment and a large range of Linux applications for the ARM
architecture.  We use the JamVM \cite{jamvm01} implementation of Sun
Microsystem's virtual machine in order to run the DSP. The Gumstix
communicates with the YSI Sonde via an RS232 serial port using a
null-modem cable since both the Sonde and the Gumstix are DTE devices.
%The serial port configuration is:

%\begin{table}[!htb]
%\caption{\label{tab:RS232_Config}RS232 Serial Port Configuration}
%\centering
%\begin{tabular}{l l}
%\hline
%Baud Rate&9600\\
%Data Bit&8\\
%Parity&None\\
%Handshake&None\\
%\hline
%\end{tabular}
%\end{table}

%\begin{table}
%\caption{\label{TAB_RS232_Pinout}RS232 Serial Pinout}
%\centering
%\begin{tabular} {|l|l|l|}
%\hline
%Wire Color	&Pin Description        &DB-9\\ 
%\hline
%Yellow 		&RS232 TX    &2\\
%Orange		&RS232 RX    &3\\  
%Green  		&Alarm       &----\\ 
%Grey			&RTS         &----\\
%Blue  		&CTS         &----\\
%Red			&+ 12V DC    &9\\
%Black			&GND         &5\\  
%Purple		&SDI-12      &----\\ 
%Bare			&Shield      &----\\
%\hline
%\end{tabular}
%\end{table}

We make use of the cellular phone network to establish a communication
link with the backend server. The cellular phone network has the
advantages of reaching well beyond telemetry options such as wireless
networks and being a significantly lower cost solution than satellite
communications. We make use of the Huawei E220. The E220 is a USB
modem that supports HSDPA, UMTS, and EDGE packet data services at
maximum transmission rates of 3.6Mbps, 384kbps, and 236.8kbps
respectively.  HSDPA and UMTS operate at 2100MHz while GSM, GPRS, and
EDGE operate at 900, 1800, and 1900MHz. The modem connects to the
Gumstix motherboard via a mini USB interface (supporting USB 2.0).The
Gumstix communicates with the modem using the PPP protocol and
accesses the cellular network using a data plan from AT\&T.

@@

Our implementation of the DSP is based on the Open Source
implementation of OSGi called Knopflerfish \cite{knopflerfish01}.
Since Sun Microsystems does not support its JDK for the Gumstix, we
use JamVM \cite{jamvm01} as the Java virtual machine implementation
and GNU Classpath \cite{classpath01} for the runtime libraries. We
have implemented various DSP Components that implement the
functionality needed for NetBEAMS. On the far left of Figure
\ref{FIG_NETBEAMS} is the actual sensor; in our case the
aforementioned YSI Sonde. The YSI Sonde DP (1) uses the RS232 standard
to read the physical measurements from the device over a serial port.
This component converts the physical measurements into a DSP-specific
message. Figure \ref{FIG_DSP_YSI_DATA} depicts the YSI Sonde's
specific message format which includes each measurement data point,
described in Table \ref{TAB_SONDE_MEASUREMENTS}.

\begin{figure}[!htb]
\centering
\epsfig{file=dsp-message-body-ysi, width=8cm, height=7cm}
\caption{\label{FIG_DSP_YSI_DATA} YSI message format.}
\end{figure}

The SondeDataContainerType defines the payload of messages generated
by the YSI Sonde DP. As all DSP messages, it is derived from the
AbstractMessageContent. The payload mainly consists of the
SondeDataType that defines all the sensor data that the YSI Sonde is
capable of producing. The YSI Sonde DP uses the RS232 to query the
physical device for its data in regular intervals. The measurements
are then converted into a DSP message and forwarded to the DSP Broker
from there they are forwarded to the remote communication component
described in the next section.  The YSI Sonde DP can also be
configured via special DSP messages, to direct the Sonde to change its
sampling frequency.  These property messages originate from a web
management interface (component (6) in Figure \ref{FIG_NETBEAMS}).

@@

All messages sent by the YSI Sonde DP need to be uploaded to a remote
database. Since the DSP itself can only handle communication within
its local DPs and DCs, we have created a matching pair of DC and DP
that can handle remote communication (components (2) and (3) in Figure
\ref{FIG_NETBEAMS}). As explained earlier, we make use of the cellular
phone network to upload sensor data to a remote backend server. While
the cellular phone network has sufficient coverage for more
environmental sensor networks, it is not cost effective to keep a data
link permanently open. The remote communiction components can be
configured to upload buffered sensor data only in certain time
intervals.Messages are transported via HTTP \cite{RFC2068}. The
matching DP/DC pair for remote communication has the following
properties:

\begin{itemize}
\item DSP Wire Transport Client: takes a collection of DSP messages
  and transmit them via HTTP. As a non-functional requirement, the
  interval in which this component will transmit messages to other
  components can be configured via the management interface.
\item DSP Wire Transport Server: is a HTTP server that receives the
  collection of DSP messages sent by the client, and piggy backs any
  messages destined for the remote client that may have queued up in
  the meantime.
\end{itemize}

The HTTP request and response are used to upload and download any
messages that have been queued at the respective end of the
communication link. The client and server component are capable of
buffering messages until they are sent to the remote destination
during the next communication cycle. Once messages have been
transported to their final destinations, they are unmarshalled and
passed on to the local DSP Broker for further dispatching. In this
regard the client and server are completely symmetrical. Messages are
dispatched with the help of the matcher to their final destinations.
